---
title: "Information on MetaWards"
author: "TJ McKinley"
date: "09/06/2020"
output: html_document
---

This is a short document (not really a vignette) describing some key aspects of MetaWards that are pertinent to this UQ attempt. 

# Model description

Firstly, the current model looks like:

```{r, echo = FALSE}
library(DiagrammeR)
grViz("digraph {
  
graph[layout = dot, rankdir = LR]

node [shape = rectangle]

S
E1 [label = 'E@_{1}']
E2 [label = 'E@_{2}']
I1 [label = 'I@_{1}']
I2 [label = 'I@_{2}']
R

S -> E1 -> E2 -> I1 -> I2 -> R
}", height = '100%', width = '100%')
```

The current prior distributions are:
\begin{align}
    \mbox{mean incubation period} &\sim U(4, 6)\\
    \mbox{mean infectious period} &\sim U(2, 4)\\
    \mbox{$R_0$} &\sim U(2.5, 4)\\
    \mbox{lockdown 1 scale} &\sim U(0, 1)\\
    \mbox{lockdown 2 scale} &\sim U(0, 1)
\end{align}
Designs generated in $(0, 1)$ or $(-1, 1)$ spaces can be easily transformed into the spaces defined above and visa-versa. Once this is done, the points on the prior space must be converted to MetaWards `beta` and `progress` parameters. We use the following transformations.

## `beta` parameters

Since $I_1$ and $I_2$ are both infectious states, they have the same $\beta$ (transmission parameter). The states are indexed from 0 in MetaWards, so $\beta_2$ is the transmission parameter from the state $I_1$ and $\beta_3$ from $I_2$. These can be derived from $R_0$ and the mean infectious period as:
\begin{align}
    \beta_2 &= \frac{R_0}{\mbox{mean infectious period}}\\
    \beta_3 &= \beta_2.
\end{align}

## `progress` parameters

These are slightly trickier, because they are not *rates*, but rather are discrete-time (daily) probabilities of transition for individuals out of each class. MetaWards uses the $E_1$ and $I_1$ classes as placeholders in order to efficiently count *incidence* of new infections and infectives. Individuals automatically stay in the $E_1$ and $I_1$ classes for exactly one day, before progressing automatically to the $E_2$ and $I_2$ classes respectively. Thus:
\begin{align}
    \mbox{progress}_0 &= 1\\
    \mbox{progress}_2 &= 1.
\end{align}
(Here $\mbox{progress}_0$ is the progression probability from the $E_1 \to E_2$ class, since progression out of the $S$ class is driven by the $\beta$ terms.) Epidemiologically this has the effect of forcing the incubation period to be *at least one day*. This is adjusted for (a bit) by removing one day from the incubation and infectious periods before calculating the other progress rates. Hence:
\begin{align}
    \mbox{progress}_1 &= 1 - \exp\left(-\frac{1}{\mbox{incubation period} - 1}\right)\\
    \mbox{progress}_3 &= 1 - \exp\left(-\frac{1}{\mbox{infectious period} - 1}\right).
\end{align}
Finally, the lockdown parameters scale the force-of-infection terms (essentially the $\beta$ rates above). These are both probabilities, so don't need any additional transformations.

## Code

A short piece of R code to do the transformations from the *prior* space to the *disease* inputs for MetaWards is:

```{r}
## function to convert samples to correct format for MetaWards
convertToMetaWards <- function(pars, repeats) {
    disease <- tibble(`beta[2]` = pars$r_zero / pars$infectious_time)
    disease$`beta[3]` <- disease$`beta[2]`
    disease$`progress[1]` <- 1 - exp(-(1 / (pars$incubation_time - 1)))
    disease$`progress[2]` <- 1
    disease$`progress[3]` <- 1 - exp(-(1 / (pars$infectious_time - 1)))
    disease$`.scale_rate[1]` <- pars$lock_1_restrict
    disease$`.scale_rate[2]` <- pars$lock_2_release
    disease$repeats <- repeats
    disease$par <- pars$par
    disease
}
```
where `pars` is an $(n \times 5)$ `data.frame` with each row corresponding to a design point, and columns called: `infectious_time`, `incubation_time`, `r_zero`, `lock_1_restrict` and `lock_1_release` (where the latter two are the scalings for full lockdown, and then slight lockdown release). The additional `repeats` argument adds the number of repeats to the matrix in the correct format for MetaWards.

The UQ4Covid tools automatically does this for a given design specified in $(-1, 1)$ space. **[EXAMPLE BELOW]**.

# Extended model description

We ideally want a model that incorporates moves to hospital and ICU. Here each class is split into two components, e.g. $E_1$ and $E_2$. This is so that MetaWards can calulate new movements into each class easily. Individuals progress out of the first of these classes afetr exactly one day with probability 1. This means that the time spent in each $E$, $I$, $A$ or $H$ class is *at least one day*. Here:

* $E$: infected but not infectious;
* $A$: asymptomatic and infectious;
* $I$: symptomatic and infectious;
* $H$: hospitalised and infectious;
* $R$: recovered and immune;
* $D$: died.

```{r, echo = FALSE}
library(DiagrammeR)
grViz("digraph {
  
graph[layout = dot, rankdir = LR]

node [shape = rectangle]

S
E1 [label = 'E@_{1}']
E2 [label = 'E@_{2}']
I1 [label = 'I@_{1}']
I2 [label = 'I@_{2}']
A1 [label = 'A@_{1}']
A2 [label = 'A@_{2}']
H1 [label = 'H@_{1}']
H2 [label = 'H@_{2}']
D
R

S -> E1 -> E2 -> {A1, I1}
A1 -> A2 -> R
I1 -> I2
I2 -> {H1, D, R}
H1 -> H2 -> {D, R}

subgraph {
  rank = same; I1; A1;
}
subgraph {
  rank = same; I2; A2;
}

}", height = '100%', width = '100%')
```

Parameters:

* $p_{SE_1}$ driven by $\beta$ parameters, defined by $R_0$ and length of infectious period (see below);
* $p_{E_1E_2} = 1$ (probability);
* $p_{E_2A_1} = p_{EA}\left(1 - e^{-\gamma^*_{E}}\right)$ where $\gamma^*_E = \frac{1}{T_E - 1}$ with $T_E$ the mean incubation period;
* $p_{E_2I_1} = \left(1 - p_{EA}\right)\left(1 - e^{-\gamma^*_{E}}\right)$;
* $p_{A_1A_2} = 1$ (probability);
* $p_{A_2R} = 1 - e^{-\gamma^*_I}$ where $\gamma^*_I = \frac{1}{T_I - 1}$ with $T_I$ the mean infectious period;
* $p_{I_1I_2} = 1$ (probability);
* $p_{I_2H_1} = p_{IH}\left(1 - e^{-\gamma^*_{I}}\right)$ where $\delta^*_I = \frac{1}{T_I - 1}$ with $T_I$ the mean infectious period;
* $p_{I_2R} = p_{IR}\left(1 - e^{-\gamma^*_{I}}\right)$;
* $p_{I_2D} = \left(1 - p_{IH} - p_{IR}\right)\left(1 - e^{-\gamma^*_{I}}\right)$;
* $p_{H_1H_2} = 1$ (probability);
* $p_{H_2D} = p_{HD}\left(1 - e^{-\gamma^*_{H}}\right)$ where $\gamma^*_H = \frac{1}{T_H - 1}$ with $T_H$ the mean length of a hospital stay;
* $p_{H_2R} = \left(1 - p_{HD}\right)\left(1 - e^{-\gamma^*_{H}}\right)$.

Lockdown can be controlled either by constraining movements over a certain distance, or by scaling the $\beta$ parameters. We do the latter at the moment.

The current prior distributions are:
\begin{align}
    \mbox{mean incubation period} &\sim U(4, 6)\\
    \mbox{mean infectious period} &\sim U(2, 4)\\
    \mbox{$R_0$} &\sim U(2.5, 4)\\
    \mbox{lockdown 1 scale} &\sim U(0, 1)\\
    \mbox{lockdown 2 scale} &\sim U(0, 1)
\end{align}

We need: **prior on length of hospital stay** (perhaps chat to Rob Challen)? We could use $U(0, 1)$ priors on many of the other parameters.

Designs generated in $(0, 1)$ or $(-1, 1)$ spaces can be easily transformed into the spaces defined above and visa-versa. Once this is done, the points on the prior space must be converted to MetaWards `beta` and `progress` parameters. We use the following transformations.



