---
title: "Information on MetaWards"
author: "TJ McKinley"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

This is a short document (not really a vignette) describing some key aspects of MetaWards that are pertinent to this UQ attempt. 

**THIS IS NOT COMPLETED YET, BUT GIVES SOME IDEAS ABOUT HOW THE MODEL IS EVOLVING.**

# Model description

We ideally want a model that incorporates moves to hospital and ICU. Here each class is split into two components, e.g. $E_1$ and $E_2$. This is so that MetaWards can calulate new movements into each class easily. Individuals progress out of the first of these classes after exactly one day with probability 1. This means that the time spent in each $E$, $I$, $A$ or $H$ class is *at least one day*. Here:

* $E$: infected but not infectious;
* $A$: asymptomatic and infectious;
* $I$: symptomatic and infectious;
* $H$: hospitalised and infectious;
* $C$: in critical care (ICU) and infectious;
* $R$: recovered and immune;
* $D$: died.

```{r, echo = FALSE}
library(DiagrammeR)
grViz("digraph {
  
graph[layout = dot, rankdir = LR]

node [shape = rectangle]

S
E1 [label = 'E@_{1}']
E2 [label = 'E@_{2}']
I1 [label = 'I@_{1}']
I2 [label = 'I@_{2}']
A1 [label = 'A@_{1}']
A2 [label = 'A@_{2}']
H1 [label = 'H@_{1}']
H2 [label = 'H@_{2}']
C1 [label = 'C@_{1}']
C2 [label = 'C@_{2}']
D
R

S -> E1 -> E2 -> {A1, I1}
A1 -> A2 -> R
I1 -> I2
I2 -> {H1, D, R}
H1 -> H2 -> {C1, D, R}
C1 -> C2 -> {D, R}

subgraph {
  rank = same; I1; A1;
}
subgraph {
  rank = same; I2; A2;
}

}", height = '100%', width = '100%')
```

## MetaWards setup

The model structure above allows for different progression pathways. MetaWards deals with this by assigning individuals to different "demographics", where each demographic can be used to represent different progression pathways. There are also `mover` functions that can be used to move individuals between demographics, and `mixer` functions that scale the force-of-infection (FOI) terms between the different demographics (explained later). There are currently seven pathways, which will be described below, but can be summarised as:

* `SEAR`: asymptomatic infections, always recover.
* `SEIR`: symptomatic infections, leading to recovery.
* `SEID`: symptomatic infections, leading to death.
* `SEIHR`: symptomatic infections, leading to hospital and then recovery.
* `SEIHD`: symptomatic infections, leading to hospital and then death.
* `SEIHCR`: symptomatic infections, leading to hospital, then critical care (ICU) and then recovery.
* `SEIHCD`: symptomatic infections, leading to hospital, then critical care (ICU) and then death.

In practice, we can summarise these through four demographics, and then `move` functions that conditionally move individuals through the seven pathways. First, we set up the `demographics.json` file, which should look something like:


```{js, code = readLines("data/MetaWards/demographics.json"), eval = FALSE}
```

Individuals in the `genpop` demographic move through the $SEIR$ and $SEID$ pathways; the `asymp` demographic moves through the `SEAR` pathway; some individuals in the `genpop` demographic can be moved to the `hospital` demographic, and thus follow the `SEIHR` and `SEIHD` pathways; and finally, some individuals in the `hospital` demographic can be moved to the `critical` demographic, and thus follow the `SEIHCR` and `SEIHCD` pathways.

We want the proportions stored in `work_ratios` and `play_ratios` to be variable parameters in the model. Therefore we want to specify proportions of individuals in each pathway such that $\sum_{i \in \mathcal{P}} p_i = 1$, where $\mathcal{P} = \{GP, A, H, C\}$.

> Here we assume the same ratios in the `work` and `play` populations, but in practice these could be changed.

### Parameters

Once the demographics have been setup, we can then control progress through the different states as:

* $p_{SE_1}$ driven by $\beta$ parameters, defined by $R_0$ and length of infectious period (see [here](#code));
* $p_{E_1E_2} = 1$;
* $p_{E_2A_1} = p_{EA}\left(1 - e^{-\gamma^*_{E}}\right)$ where $\gamma^*_E = \frac{1}{T_E - 1}$ with $T_E$ the mean incubation period;
* $p_{A_1A_2} = 1$;
* $p_{A_2R} = 1 - e^{-\gamma^*_I}$ where $\gamma^*_I = \frac{1}{T_I - 1}$ with $T_I$ the mean infectious period;
* $p_{E_2I_1} = \left(1 - p_{EA}\right)\left(1 - e^{-\gamma^*_{E}}\right)$;
* $p_{I_1I_2} = 1$;
* $p_{I_2H_1} = p_{IH}\left(1 - e^{-\gamma^*_{I}}\right)$ where $\gamma^*_I = \frac{1}{T_I - 1}$;
* $p_{I_2R} = p_{IR}\left(1 - e^{-\gamma^*_{I}}\right)$;
* $p_{I_2D} = \left(1 - p_{IH} - p_{IR}\right)\left(1 - e^{-\gamma^*_{I}}\right)$;
* $p_{H_1H_2} = 1$;
* $p_{H_2C_1} = p_{HC}\left(1 - e^{-\gamma^*_{HC}}\right)$ where $\gamma^*_{HC} = \frac{1}{T_H - T_C - 1}$ with $T_H$ the mean length of a hospital stay and $T_C$ is the mean length of time spent in ICU;
* $p_{H_2R} = p_{HR}\left(1 - e^{-\gamma^*_{H}}\right)$ where $\gamma^*_H = \frac{1}{T_H - 1}$;
* $p_{H_2D} = \left(1 - p_{HC} - p_{HR}\right)\left(1 - e^{-\gamma^*_{H}}\right)$;
* $p_{C_1C_2} = 1$;
* $p_{C_2R} = p_{CR}\left(1 - e^{-\gamma^*_{C}}\right)$ where $\gamma^*_C = \frac{1}{T_C - 1}$ with $T_C$ is the mean length of time spent in ICU;
* $p_{C_2D} = \left(1 - p_{CR}\right)\left(1 - e^{-\gamma^*_{C}}\right)$.

Lockdown can be controlled either by constraining movements over a certain distance, or by scaling the $\beta$ parameters. We do the latter at the moment.

### Disease `.json` file

All pathways have an $SEI(R/D)$ structure, so we can set up an `ncov.json` file for the overall disease:

```{js, code = readLines("data/MetaWards/ncov.json"), eval = FALSE}
```

> **Note:** We would want asymptomatics to be less infectious than symptomatics, so could amend $\beta$, or set `contrib_foi`?

### Moving functions

We need moving functions to move individuals from $E_2 \to A_1$, $E_2 \to I_1$, $I_2 \to H_1$ and $H_2 \to C_1$, and thus we need a custom `move` function. Create a file `move_pathways.py` containing:

```{python, code = readLines("data/MetaWards/move_pathways.py"), eval = FALSE}
```

### Interaction matrices

The **interaction matrix** scales the FOI that different demographics have on other demographics. We need to store this in a file called `mix_pathways.py`. Since most of the classes are already infected, it's only the first row that has to contain non-zero entries. Here we've assumed zero transmission from hospitals to the general population, and used a `GP_A` parameter to scale the impact of aymptomatics on infection in the general population.

```{python, code = readLines("data/MetaWards/mix_pathways.py"), eval = FALSE}
```

### Code {#code}

To run designs, we need to generate a `disease.csv` file containing different parameters to use for different runs. For consistency, we will define three spaces:

* *input* space: this relates to the parameters ranges (defined below);
* *design* space: this will usually be in $(0, 1)$ or $(-1, 1)$ space;
* *disease* space: this relates to parameters that are fed into MetaWards.

The *input* and *design* spaces are fairly trivial to convert between, but some more work has to be done to convert between the *input* space and the *disease* space.

The current *input* parameter ranges are:
\begin{align}
    \mbox{$R_0$}&: (2.5, 4)\\
    \mbox{scale asymptomatics}&: (0, 1)\\
    \mbox{mean incubation period ($T_E$)}&: (4, 6)\\
    \mbox{mean infectious period ($T_I$)}&: (2, 4)\\
    \mbox{mean hospital stay ($T_H$)}&: (4, 12)\\
    \mbox{proportion of hospital time spent in critical care ($p_{tC}$)}&: (0, 1)\\
    \mbox{lockdown 1 restrict}&: (0, 1)\\
    \mbox{lockdown 2 release}&: (0, 1)\\
    p_{EA}&: (0, 1)\\
    p_{IH}&: (0, 1)\\
    p_{IR}&: (0, 1) \quad \mbox{such that}~\sum_{i \in \{H, R\}} p_{Ii} < 1\\
    p_{HC}&: (0, 1)\\
    p_{HR}&: (0, 1) \quad \mbox{such that}~\sum_{i \in \{C, R\}} p_{Hi} < 1\\
    p_{CR}&: (0, 1)
\end{align}
We have that the time in critical care, $T_C$, must be a proportion of the time spent in hospital, $T_H$, and thus
$$
    T_C = p_{tC}T_H.
$$

We need: **ranges for the length of hospital stay** (perhaps chat to Rob Challen)? Current hospital rough guides taken from [https://www.medrxiv.org/content/10.1101/2020.04.23.20076042v1.full.pdf](https://www.medrxiv.org/content/10.1101/2020.04.23.20076042v1.full.pdf). (I've stuck in IQR for comparison, but clearly we need to think about this more.)

To ensure orthogonality, we could reparameterise such that:
\begin{align}
    p_{IR} &= \left(1 - p_{IH}\right) p^\prime_{IR}\\
    p_{HR} &= \left(1 - p_{HC}\right) p^\prime_{HR}
\end{align}
and thus we have parameter ranges:
\begin{align}
    \mbox{$R_0$}&: (2.5, 4)\\
    \mbox{scale asymptomatics}&: (0, 1)\\
    \mbox{mean incubation period ($T_E$)}&: (4, 6)\\
    \mbox{mean infectious period ($T_I$)}&: (2, 4)\\
    \mbox{mean hospital stay ($T_H$)}&: (4, 12)\\
    \mbox{proportion of hospital time spent in critical care ($p_{tC}$)}&: (0, 1)\\
    \mbox{lockdown 1 restrict}&: (0, 1)\\
    \mbox{lockdown 2 release}&: (0, 1)\\
    p_{EA}&: (0, 1)\\
    p_{IH}&: (0, 1)\\
    p^\prime_{IR}&: (0, 1)\\
    p_{HC}&: (0, 1)\\
    p^\prime_{HR}&: (0, 1)\\
    p_{CR}&: (0, 1)
\end{align}
and this has removed the sum-to-one-constraints.

In R we can set up the *input* parameter ranges as follows:

```{r}
## set up parameter ranges
parRanges <- data.frame(
    parameter = c("r_zero", "scale_asymp", "incubation_time", "infectious_time", "hospital_time",
                  "ptC", "lock_1_restrict", "lock_2_release",
                  "pEA", "pIH", "pIRprime", "pHC", "pHRprime", "pCR"),
    lower = c(2.5, 0, 4, 2, 4, rep(0, 9)),
    upper = c(4, 1, 6, 4, 12, rep(1, 9)),
    stringsAsFactors = FALSE
)
```

Firstly we want a function to convert between the *design* and *input* spaces. A short R function to do this is given below. This requires a `design` data frame, with columns denoting each *input* parameter in `parRanges` and rows corresponding to design points, and a `parRanges` data frame (defined above). We use the `scale` argument to define whether the design is on the $(0, 1)$ (`scale = "zero_one"`) or $(-1, 1)$ (`scale = "negone_one"`) space.

```{r, echo = FALSE}
library(knitr)
read_chunk("data/MetaWards/dataTools.R")
```

```{r, convertDesignToInput}
```

Once we have done this, we need to transform from the *input* space to the *disease* space for MetaWards. An R function is given below. This requires an `input` data frame, with columns denoting each *input* parameter and rows corresponding to each input points, and a number of `repeats`.


```{r, convertInputToDisease}
```

> **TODO: add attenuated infection rate for asymptomatics.**

> **QUESTION**: do we want to set $T_H$ to be a subset of $T_I$, or in addition to $T_I$? How do we specify ranges?

> **QUESTION**: how do we scale FOI from hospital patients to the other populations?

As an example of a quick LHS design for five design points and five replicates:

```{r, code = readLines("data/MetaWards/convertDesign.R"), eval = FALSE}
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
tempdir <- getwd()
setwd("./data/MetaWards/")
source("convertDesign.R")
setwd(tempdir)
```

> **TODO: might want some form of folder clearup or suchlike?**



