---
title: "Running Our Metawards implementation for yourself (for dummies)"
author: "Danny Williamson"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

When performing UQ, it can hardly be overstated how useful it can be to run the model you are analysing for yourself. Obviously, if you can, as with Metawards, this comes with its own problems and dangers, and you might run the model in such a way as not to get anything useful out. Thanks to the work of our team (TJ McKinley and Chris Fenton in particular), and with help from Metawards developer Chris Woods, we have a model structure [described here](https://uq4covid.github.io/vignettes/data/MetaWards/vignette/MetaWards) and a stable version of the code for running it for UQ and extracting the data for yourselves! 

So why am I writing a "for dummies" guide to using it and not Chris/TJ? Well, because I qualify as a member of the target audience! Today, I went through TJ's instructions, failed in several places and figured out everything I needed to do to run it, so I figured I would pass that knowledge on.

## Instructions for Linux/Mac

As Mac uses unix and I am on a mac, the instructions are identical. There will need to be tweaks to the below for windows users, but what is here should be enough to get you started.

### Install Metawards

The easiest thing to do is to open a terminal and type
```{unix, eval=FALSE}
pip3 install metawards
```

I used pip3, you may have Python 3 as default and can use pip. The absolute best thing to do is go [here](https://metawards.org/install.html) and follow the instructions. They are really clear. For this part it doesnt matter what directory you are in.

Now, you need to head to a directory where you want the Metawards data to live. If not familar with terminal navigation, I used
```{unix, eval=FALSE}
cd Dropbox/BayesExeter
```

and you can check if you are happy with whats in the directory with
```{unix, eval=FALSE}
ls
```

Next install the metawards data files with 
```{unix, eval=FALSE}
git clone https://github.com/metawards/MetaWardsData
```

### Get the uq4covid version

Now, we get our version of the model, which consists of a number of files needed to do the running, designing and extracting. These are on the `uq4covid.io` repository [here](https://github.com/UQ4covid/uq4covid.github.io), which you should clone onto your machine or pull if you have it already.

Once you have it, go to directory `vignettes/data/MetaWards` and there is everything you need. **To ensure we keep the repo version clean, it's best to copy this folder out to somewhere else.** This guide is going to use Rstudio to do most things, so head to this folder in Rstudio and set it as your working directory.

### Generating a new design

The file `convertDesign.R` creates new designs to your specification. To run a custom design, you need to open it and change a few things. I'll highlight the blocks you need to look at now.

First the parameter ranges look like this
```{r, eval=FALSE}
parRanges <- data.frame(
    parameter = c("r_zero", "incubation_time", "infectious_time", "hospital_time",
                  "critical_time", "lock_1_restrict", "lock_2_release",
                  "pEA", "pIH", "pIRprime", "pHC", "pHRprime", "pCR", 
                  "GP_A", "GP_H", "GP_C"),
    lower = c(2.5, 4, 2, 4, 4, rep(0, 11)),
    upper = c(4, 6, 4, 12, 12, rep(1, 11)),
    stringsAsFactors = FALSE
)
```

You can alter `lower` and `upper` for any of the parameters (or even fix some if you like) here. See the model description linked at the top to see what they all do.

Next, generate your own design. The default is a test 5 member LHC and is in the code like this
```{r, eval=FALSE}
design <- randomLHS(5, nrow(parRanges))
colnames(design) <- parRanges$parameter
design <- as_tibble(design)
```

Obviously, change the design to something of your own choosing (remembering at all times how bad random Latin Hypercubes can be), replacing the first of these 3 lines with your own.

The next 2 lines are super important
```{r, eval = FALSE}
design$output <- ensembleIDGen(ensembleID = "Ens0", nrow(design))
design$repeats <- 2
```

The second gives the number of repeats per run so is fairly obvious. Our early experiments showed that at least 20 might be needed for some good techniques to work. The output argument is very important. This is a unique identifier for the ensemble. `"Ens0"` is the first test ensemble we've run. `"Ens1"` will be the next. Soon there will be `"Ens9"`, `"Ensa"`, `"Ensb"` and so on. We use "Ens" as Metawards does funny things if you use numbers or some single letters. Ideally, you would let us know you are generating a big ensemble with the next ID for putting on the git page. This identifier will ultimately be responsible for identifying your ensemble from others, and for matching inputs to outputs during extraction. Change it now.

One you have made those changes, run the whole script and you are ready to run metawards!

### Running Metawards

There are 2 simple steps to this. First, you need to edit the file `runscript.sh`. Simply open it with xcode, and alter the line
```{unix, eval=FALSE}
export METAWARDSDATA=$HOME/Dropbox/BayesExeter/MetaWardsData
```
so that the direcroty is the place you cloned the Metawards data above. Now it's run time! The default is to run for 100 days, but you will probably want longer to look at interesting things. The final part of the long line calling Metawards handles this
```{unix, eval=FALSE}
--nsteps 100
```

and you can alter it in obvious ways.

Open up a terminal and navigate via `cd` to the folder where your `runscript.sh` is. Once there type
```{unix, eval=FALSE}
./runscript.sh
```

and let the fun begin. Even this small test ensemble took about 10 mins on my machine, and it runs on all the cores, so don't expect to be able to do a tonne whilst it's running. TJ had a bigger job run all night.

### Extracting data

Eventually the fans will cool down and the simulations will be complete. Now you will want to get at some of the outputs. The `extractOutput.R` file lets you do this. Open this in RStudio.

If you have all the required libraries, you can just run the whole thing and, by default, you will get the total number of infections in hospitals per ward. There is a lot you can do to customise, and I think TJ will write a vignette with more information. To get you started, here are some things you can change. 

The chunk
```{r, eval=FALSE}
hospital <- tbl(con, "hospital_totals")
        hosp_db <- filter(hospital, day <= 100) %>%
            select(ward, stage_2) %>%
            group_by(ward) %>%
            summarise(cH = sum(stage_2))
        ## collect outcome of query
        hosp <- collect(hosp_db)
```

is taking the database `con` created from the raw outputs, is selecting those from the `hospital` demographic, and summing the numbers at stage 2. Some reference to our model setup is required to see what to do, so maybe head there to see what the stages and demographics refer to in the diagram. We have 4 demographics:
`genpop` , `hospital` , `critical`, and `asymp` (so you can substitute for any of these). There are 6 stages indexed 0-5 (so above we have stage 2). The diagram in TJ's instruction shows where they are, but there are 2 things to note. Stages 0 and 2 count the new incidents of entry into that demographic *that day*. Stages 1 and 3 count the total incidents in that demographic that day *including the new ones from 0 and 1 respectively*. It's a postprocessing trick, meaning that we get the most informative data each time. Stage 4 is recovered and stage 5 is death.

Having extracted data like this in tibble form, it is ready to be processed for building whatever emulators you like.

### Sundries

All of the files for running Metawards are in the different folders in our repo (that you will have cloned onto your machine), so you are free to change things for your own investigations. In the future, when we decide on different experiments we want to do, we will alter these files, but still be able to retain the same designing, running and processing experience.

Happy UQing!
