## MetaWards Model Structure {#model}

> **In order to get the model structure and outputs described previously, we have to play a few tricks in MetaWards. The detail below corresponds to how we have implemented the current model, and requires additional stages to ensure we get the correct transitions and outputs.**

Here most classes are split into two components, e.g. $E_1$ and $E_2$. This is so that MetaWards can calculate new movements into each class easily, but are otherwise epidemiologically identical. Here:

* $E$: infected but not infectious;
* $A$: asymptomatic and infectious;
* $I$: symptomatic and infectious;
* $H$: hospitalised and infectious;
* $C$: in critical care (ICU) and infectious;
* $R$: recovered and immune;
* $D$: died.

```{r, echo = FALSE}
library(DiagrammeR)
grViz("digraph model {
    graph[layout = dot, rankdir = LR]
    node [shape = rectangle]
    
    S
    E1 [label = 'E@_{1}']; E2 [label = 'E@_{2}']
    I1 [label = 'I@_{1}']; I2 [label = 'I@_{2}']
    A1 [label = 'A@_{1}']; A2 [label = 'A@_{2}']
    H1 [label = 'H@_{1}']; H2 [label = 'H@_{2}']
    C1 [label = 'C@_{1}']; C2 [label = 'C@_{2}']
    DI [label = 'D@_{I}']; DH [label = 'D@_{H}']; DC [label = 'D@_{C}']
    RI [label = 'R@_{I}']; RH [label = 'R@_{H}']; RC [label = 'R@_{C}']; RA [label = 'R@_{A}']
    St00 [style = 'invis']; St0 [label = 'Stage 0']; St1 [label = 'Stage 1']; 
    St2 [label = 'Stage 2']; St3 [label = 'Stage 3']; St4 [label = 'Stage 4']; St5 [label = 'Stage 5']
    genpop [label = 'Gen. Pop.']
    hospital [label = 'Hospital Patient']
    critical [label = 'Critical Care']
    asymp [label = 'Asymptomatics']
    
    St00 -> St0 -> St1 -> St2 -> St3 -> St4 -> St5 [style = 'invis']
    S -> E1 -> E2 -> I1 -> I2 
    I2 -> RI
    I2 -> DI [weight = 0]
    I2 -> H1 -> H2 -> C1 -> C2 
    C2 -> RC 
    C2 -> DC [weight = 0]
    H2 -> RH 
    H2 -> DH [weight = 0] 
    RI -> DI [style = 'invis']
    RH -> DH [style = 'invis']
    RC -> DC [style = 'invis']
    E2 -> A1 [weight = 0]
    A1 -> A2 -> RA 
    genpop -> S [style = 'invis']
    hospital -> H1 [style = 'invis']
    critical -> C1[style = 'invis']
    asymp -> A1 [style = 'invis']
    subgraph {
      rank = same; St00; genpop; hospital; critical; asymp;
    }
    subgraph {
      rank = same; E1; St0;
    }
    subgraph {
      rank = same; E2; St1;
    }
    subgraph cluster {
      subgraph {
        rank = same; I1; H1; C1; A1; St2;
      }
      subgraph {
        rank = same; I2; H2; C2; A2; St3;
      }
      style = filled;
      color = lightgrey;
      label = 'Infectious Classes'
    }
    subgraph {
      rank = same; RA, RI, RC; RH; St4;
    }
    subgraph {
      rank = same; DI; DH; DC; St5;
    }
    
}", height = '100%', width = '100%')
```

### MetaWards setup

The model structure above allows for different progression pathways. MetaWards deals with this by assigning individuals to different "demographics", where each demographic can be used to represent different progression pathways. There are also `mover` functions that can be used to move individuals between demographics, and `mixer` functions that scale the force-of-infection (FOI) terms between the different demographics (explained later). There are currently seven pathways, which will be described below, but can be summarised as:

* `SEAR`: asymptomatic infections, always recover.
* `SEIR`: symptomatic infections, leading to recovery.
* `SEID`: symptomatic infections, leading to death.
* `SEIHR`: symptomatic infections, leading to hospital and then recovery.
* `SEIHD`: symptomatic infections, leading to hospital and then death.
* `SEIHCR`: symptomatic infections, leading to hospital, then critical care (ICU) and then recovery.
* `SEIHCD`: symptomatic infections, leading to hospital, then critical care (ICU) and then death.

In practice, we can summarise these through four demographics, and then `move` functions that conditionally move individuals through the seven pathways. First, we set up the `demographics.json` file, which should look something like:


```{js, code = readLines("../model_code/demographics.json"), eval = FALSE}
```

Individuals in the `genpop` demographic move through the $SEIR$ and $SEID$ pathways; the `asymp` demographic moves through the `SEAR` pathway; some individuals in the `genpop` demographic can be moved to the `hospital` demographic, and thus follow the `SEIHR` and `SEIHD` pathways; and finally, some individuals in the `hospital` demographic can be moved to the `critical` demographic, and thus follow the `SEIHCR` and `SEIHCD` pathways.

We need to specify initial proportions of individuals in each pathway such that $\sum_{i \in \mathcal{P}} p_i = 1$, where $\mathcal{P} = \{GP, A, H, C\}$.

> Here we assume the same ratios in the `work` and `play` populations, but in practice these could be changed.

### Parameters {#parameters}

Once the demographics have been setup, we can then control progress through the different states as:

* $p_{SE_1}$ driven by $\beta$ parameters, defined by $R_0$ and length of infectious period (see [here](#code));
* $p_{E_1E_2} = 1$;
* $p_{E_2A_1} = p_{EA}\left(1 - e^{-\gamma_{E}}\right)$ where $\gamma_E = \frac{1}{T_E}$ with $T_E$ the mean incubation period;
* $p_{E_2I_1} = \left(1 - p_{EA}\right)\left(1 - e^{-\gamma_{E}}\right)$;
* $p_{A_1A_2} = 1$;
* $p_{A_2R} = 1 - e^{-\gamma_I}$ where $\gamma_I = \frac{1}{T_I}$ with $T_I$ the mean infectious period;
* $p_{I_1I_2} = 1$;
* $p_{I_2H_1} = p_{IH}\left(1 - e^{-\gamma_{I}}\right)$;
* $p_{I_2R} = p_{IR}\left(1 - e^{-\gamma_{I}}\right)$;
* $p_{I_2D} = \left(1 - p_{IH} - p_{IR}\right)\left(1 - e^{-\gamma_{I}}\right)$;
* $p_{H_1H_2} = 1$;
* $p_{H_2C_1} = p_{HC}\left(1 - e^{-\gamma_{H}}\right)$ where $\gamma_{H} = \frac{1}{T_H}$ with $T_H$ the mean length of a hospital stay;
* $p_{H_2R} = p_{HR}\left(1 - e^{-\gamma_{H}}\right)$;
* $p_{H_2D} = \left(1 - p_{HC} - p_{HR}\right)\left(1 - e^{-\gamma_{H}}\right)$;
* $p_{C_1C_2} = 1$;
* $p_{C_2R} = p_{CR}\left(1 - e^{-\gamma_{C}}\right)$ where $\gamma_C = \frac{1}{T_C}$ with $T_C$ is the mean length of time spent in ICU;
* $p_{C_2D} = \left(1 - p_{CR}\right)\left(1 - e^{-\gamma_{C}}\right)$.

Lockdown can be controlled either by constraining movements over a certain distance, or by scaling the $\beta$ parameters. We do the latter at the moment. The moves of the type $p_{X_1X_2} = 1$ are tricks to enable us to get the correct counts (see [below](#movements) for more details).

### Disease `.json` file {#diseasefile}

All pathways have an $SEI(R/D)$ structure, so we can set up an `ncov.json` file for the overall disease:

```{js, code = readLines("../model_code/ncov.json"), eval = FALSE}
```

We will discuss these choices in more detail in the subsequent sections, in particular the role of the `beta` parameters in the next section. Note that the `progress` parameters are all set to zero here, since all movements will be controlled via the custom `mover` function.

### Controlling movements {#movements}

MetaWards has a specific structure in terms of how it progresses movements between the stages. To get the correct splits between the pathways specified [above](#model) we do all non-infection movements through a custom `mover` function specified [below](#mover). Note that the $X_1$-type states (e.g. $E_1$, $I_1$, $H_1$ and $C_1$) are dummy states that allow us to monitor *new moves into each class at each day*. Note that in the `ncov.json` file specified [above](#diseasefile), we set all `progress` parameters to be 0. Thus, all transition probabilities other than new infections are driven by user-defined parameters that are passed to the `mover` function.

All this effort is so that we can monitor the counts and ensure the correct transitions, and depends on the order in which the events happen. For example, imagine that on day $t$ we get an infection at the end of the day, so a move from $S \to E_1$. Then the **recorded counts** for $E_1$ contain the new infections. Then, at the beginning of day $t + 1$, these individuals will be automatically moved from $E_1 \to E_2$ by the mover function, which subsequently operates using the probabilities [above](#parameters) to ensure the correct proportion of individuals progress into the $I_1$ or $A_1$ classes. The `mover` function applies movements in order, and so it is important to get the order correct. In particular we need to reverse the order of the stage movements (e.g. do movements out of the $C$ demographic *before* movements out of the $H$ demographic). This is to ensure that individuals move from $H_2 \to C_1$ say, can't then immediately move out of $C_1$. However, *within* each stage, we need to do $X_1 \to X_2$ moves first so that these individuals can progress out of $X_2$ subsequently (since the $X_1$ stages are simply dummy stages used for counting).

> **Note**: remember that $X_1$ and $X_2$ are tricks here. We want $X_1$ to record the number of **new movements into** class $X$, and $X_2$ to record the **number of individuals in** class $X$. At the moment, moves into $X_1$ on day $t$ say are not currently included in $X_2$. Epidemiologically these are treated OK, since the [`mover`](#mover) below accounts for the transitions correctly. However, we want the outputs for the $X_2$ type classes to be the **number of individuals** in that class on each day. As such, we use a custom [`extractor`](#extractor) function that adds $X_1$ to $X_2$ at the end, so that $X_1$ corresponds to **new movements into** the $X$ class, and $X_2$ is the **number** of individuals in the $X$ class.

#### Moving functions {#mover}

The implement the above ideas, create a file `move_pathways.py` containing the code below (hidden for brevity, but can be toggled on).

```{python, code = readLines("../model_code/move_pathways.py"), eval = FALSE}
```

> **Additional note**: The functions in the `mover` file operate in turn. Therefore the movement probabilities [above](#parameters) must be altered between each function, in order to get the correct proportions moved. For example, consider that we have $n$ individuals in the $I_2$ class and we want to move a proportion $p_Ip_{IH}$ from $I_2 \to H_1$, a proportion $p_Ip_{IR}$ from $I_2 \to R_I$, and a proportion $p_I\left(1 - p_{IH} - p_{IR}\right)$ from $I_2 \to D_I$, where $p_I = 1 - e^{-\gamma_I}$.
> 
> In this case the first `mover` function takes a random binomial sample from the $n$ individuals with probability $p_Ip_{IH}$ as requested, resulting in $n_{IH}$ moves. However, the second `mover` function now operates on the $n - n_{IH}$ individuals, so we need to adjust the sampling probabilities to adjust for this. Hence the second `mover` function needs to sample from the $n - n_{IH}$ individuals with probability $\frac{p_Ip_{IR}}{1 - p_Ip_{IH}}$ in order to generate the correct proportions of moves that we would expect, resulting in $n_{IR}$ moves. Similarly, the third `mover` function now operates on the $n - n_{IH} - n_{IR}$ remaining individuals, and thus we would need to adjust the sampling probability to $\frac{p_I\left(1 - p_{IH} - p_{IR}\right)}{1 - p_I\left(p_{IH} + p_{IR}\right)}$. The remaining individuals remain in $I_2$.

### Interaction matrices

The **interaction matrix** scales the FOI that different demographics have on other demographics. We need to store this in a file called `mix_pathways.py`. Since most of the classes are already infected, it's only the first row that has to contain non-zero entries. Here the `GP_A`, `GP_H` and `GP_C` parameters scale the impact of aymptomatics, hospital cases and critical care cases on the force-of-infection to the general population. All demographics other than `genpop` are already infected, so have no further force-of-infection acting on them, and thus the other parameters can be set to zero. These parameters are all $> 0$.

```{python, code = readLines("../model_code/mix_pathways.py"), eval = FALSE}
```

### Lockdown

Chris Fenton wrote a neat custom `iterator` function, that models lockdown by scaling the FOI terms for different time period,s that represent different stages of interventions. The custom iterator is:

```{python, code = readLines("../model_code/iterate.py"), eval = FALSE}
```

This requires an additional input file containing the dates of the different stages (which are fixed parameters here). This is stored in the `lockdown_states.txt` file:

```{python, code = readLines("../model_code/lockdown_states.txt"), eval = FALSE}
```

### Extractor {#extractor}

We also have a custom `extractor` function, that saves the outputs as a compressed SQL database called `stages.db.bz2`. This is important since the outputs are very large (all classes across all days and all wards, for multiple replicates of each design point). The database contains four tables:

* `genpop_totals`
* `hospital_totals`
* `asymp_totals`
* `critical_totals`

You will notice these are named after the different **demographics**. Each table contains entries for:

* `ward`, `day`, `stage_0`, `stage_1`, `stage_2`, `stage_3`, `stage_4`, `stage_5`

For the `genpop` demographic, these correspond to:

* `stage_0`: **new** infections on each day (i.e. new moves into $E$)
* `stage_1`: current infections on each day (i.e. number of individuals in $E$)
* `stage_2`: **new** infectious individuals on each day (i.e. new moves into $I$)
* `stage_3`: current infectious individuals on each day (i.e. number of individuals in $I$)
* `stage_4`: current removals on each day (i.e. number of individuals in $R$)
* `stage_5`: current deaths on each day (i.e. number of individuals in $D$)

The other demographics (`asymp`, `hospital` and `critical`) have the same structure, but with `stage_0` and `stage_1` **always 0** (since new infections are always in `genpop`), and `stage_2` and `stage_3` correspond to the new and current numbers in either the $A$, $H$ or $C$ classes according to the demographic (refer to the model structure [here](#modelS)).

### Input and output code {#code}

To run designs, we need to generate a `disease.csv` file containing different parameters to use for different runs. For consistency, we will define three spaces:

* *input* space: this relates to the parameters ranges (defined below);
* *design* space: this will usually be in $(0, 1)$ or $(-1, 1)$ space;
* *disease* space: this relates to parameters that are fed into MetaWards.

The *input* and *design* spaces are fairly trivial to convert between, but some more work has to be done to convert between the *input* space and the *disease* space.

The current *input* parameter ranges are:
\begin{align}
    \mbox{$R_0$}&: (2.5, 4)\\
    \mbox{mean incubation period ($T_E$)}&: (4, 6)\\
    \mbox{mean infectious period ($T_I$)}&: (2, 4)\\
    \mbox{mean hospital stay ($T_H$)}&: (4, 12)\\
    \mbox{mean stay in critical care ($T_C$)}&: (4, 12)\\
    \mbox{lockdown 1 restrict}&: (0, 1)\\
    \mbox{lockdown 2 release}&: (0, 1)\\
    p_{EA}&: (0, 1)\\
    p_{IH}&: (0, 1)\\
    p_{IR}&: (0, 1) \quad \mbox{such that}~\sum_{i \in \{H, R\}} p_{Ii} < 1\\
    p_{HC}&: (0, 1)\\
    p_{HR}&: (0, 1) \quad \mbox{such that}~\sum_{i \in \{C, R\}} p_{Hi} < 1\\
    p_{CR}&: (0, 1)\\
    \beta^S_{A \to GP}&: (0, 1)\\
    \beta^S_{H \to GP}&: (0, 1)\\
    \beta^S_{C \to GP}&: (0, 1)
\end{align}
(The $\beta^S_{C \to GP}$ type parameters are the scaling rates used in the interaction matrix.)

We need: **ranges for the length of hospital stay** (perhaps chat to Rob Challen)? Current hospital rough guides taken from [https://www.medrxiv.org/content/10.1101/2020.04.23.20076042v1.full.pdf](https://www.medrxiv.org/content/10.1101/2020.04.23.20076042v1.full.pdf). (I've stuck in IQR for comparison, but clearly we need to think about this more.)

To ensure orthogonality, we could reparameterise such that:
\begin{align}
    p_{IR} &= \left(1 - p_{IH}\right) p^\prime_{IR}\\
    p_{HR} &= \left(1 - p_{HC}\right) p^\prime_{HR}
\end{align}
and thus we have parameter ranges:
\begin{align}
    \mbox{$R_0$}&: (2.5, 4)\\
    \mbox{mean incubation period ($T_E$)}&: (4, 6)\\
    \mbox{mean infectious period ($T_I$)}&: (2, 4)\\
    \mbox{mean hospital stay ($T_H$)}&: (4, 12)\\
    \mbox{mean time in critical care ($T_C$)}&: (4, 12)\\
    \mbox{lockdown 1 restrict}&: (0, 1)\\
    \mbox{lockdown 2 release}&: (0, 1)\\
    p_{EA}&: (0, 1)\\
    p_{IH}&: (0, 1)\\
    p^\prime_{IR}&: (0, 1)\\
    p_{HC}&: (0, 1)\\
    p^\prime_{HR}&: (0, 1)\\
    p_{CR}&: (0, 1)\\
    \beta^S_{A \to GP}&: (0, 1)\\
    \beta^S_{H \to GP}&: (0, 1)\\
    \beta^S_{C \to GP}&: (0, 1)
\end{align}
and this has removed the sum-to-one-constraints. 

In R we can set up the *input* parameter ranges as follows:

```{r}
## set up parameter ranges
parRanges <- data.frame(
    parameter = c("r_zero", "incubation_time", "infectious_time", "hospital_time",
                  "critical_time", "lock_1_restrict", "lock_2_release",
                  "pEA", "pIH", "pIRprime", "pHC", "pHRprime", "pCR", 
                  "GP_A", "GP_H", "GP_C"),
    lower = c(2.5, 4, 2, 4, 4, rep(0, 11)),
    upper = c(4, 6, 4, 12, 12, rep(1, 11)),
    stringsAsFactors = FALSE
)
```

Firstly we want a function to convert between the *design* and *input* spaces. A short R function called `convertDesignToInput()` which does this is given below. This requires a `design` data frame with columns denoting each *input* parameter in `parRanges` and rows corresponding to design points. There should be an additional column called `output` that defines a unique identifier for each design point, and a column called `repeats` that contains the number of repeats for each design point. The `convertDesignToInput()` function also requires the `parRanges` data frame (defined above). We use the `scale` argument to define whether the design is on the $(0, 1)$ (`scale = "zero_one"`) or $(-1, 1)$ (`scale = "negone_one"`) space.

> **Note**: this function converts the probabilities correctly for the `mover` defined [above](#mover).

```{r, echo = FALSE}
library(knitr)
read_chunk("../R_tools/dataTools.R")
```

```{r, convertDesignToInput}
```

Once we have done this, we need to transform from the *input* space to the *disease* space for MetaWards. An R function is given below. This requires an `input` data frame, with columns denoting each *input* parameter and rows corresponding to each input points, a number of `repeats` and a column of unique identifiers (`output`).


```{r, convertInputToDisease}
```

Also in `dataTools.R` is a function `ensembleIDGen()` that creates unique IDs for each design point. So an example of a quick LHS design for five design points and five replicates is:

```{r, code = readLines("../convertDesign.R"), eval = FALSE}
```

This produces a file `inputs/disease.dat` that can be passed to MetaWards to run the model. The `runscript.sh` batch file provides the command line instructions needed to the model using these inputs. This is written for Linux, and should be able to be run directly from the downloaded folder. If you don't run Linux, then the file should give you an idea of how to run the model on your own system.

The outputs from MetaWards are stored in an SQL database, which can be queried through any SQLite client. Each model run produces a file called `stages.db.bz2`, which is a compressed database containing the outputs. As an example, migrate to an output folder containing `stages.db.bz2`. To access the outputs you will first have to unzip this file. I do this on the command line e.g.

```
bzip2 -dk stages.db.bz2
```

If you're happy with SQL, you can query these directly with e.g. SQLite. If you are an R user, then the `dplyr` package (or more specifically the `dbplyr` package) provides some useful R tools for querying SQL databases using `tidyverse`-type notation. More details can be found [here](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html).

As a quick example, imagine that we want to extract the **cumulative hospital cases** on say day 100. Here we will need to extract the **new hospital cases** from day 1--100, and then sum them up for each ward. Therefore we need to extract `day`, `ward` and `stage_2` from the `hospital_totals` table, and then sum them up (see [here](#extractor) for more details on how the outputs are stored). For speed, `dplyr` converts the R code to SQL, and only pulls the resulting table to R if you run the function `collect()` (see below).

```{r, eval = FALSE}
## load library
## (you might also need to install the 'RSQLite' 
## and `dbplyr` packages which 'dplyr' calls)
library(dplyr)

## establish connection to database
con <- DBI::dbConnect(RSQLite::SQLite(), "stages.db")

## connect to the 'hospital_totals' table
hospital <- tbl(con, "hospital_totals")

## stage_2 contains the new cases, so sum these
## over each ward for days 1--100 
## (NOTE: `hosp_db` contains the query not the outcome,
## so 'show_query(hosp_db)' can view this if you like)
hosp_db <- filter(hospital, day <= 100) %>%
    select(ward, stage_2) %>%
    group_by(ward) %>%
    summarise(cH = sum(stage_2))

## run query and pull to R
hosp <- collect(hosp_db)

## disconnect from database
DBI::dbDisconnect(con)
```

Now you can play with `hosp` as much as you like. Note that after you've pulled the correct outputs down, you might want to delete the `stage.db` (**NOT** the `stages.db.bz2`) file. I did this on the Unix command line using:

```
rm stage.db
```

> **Be careful**: remember that `rm` removes completely, so not to the recycle bin. However, as long as you don't remove `stages.db.bz2` then you can always recover.

You can very easily wrap these ideas into an R function that can scroll through the design IDs, extract relevant outputs and bind to the inputs. An example that you are free to edit at will can be found in the `extractOutputs.R` file in the repo. **At the current time this relies on some OS-specific code, in order to unzip the compressed SQL database files.**

```{r, echo = FALSE, message = FALSE, warning = FALSE}
tempdir <- getwd()
setwd("..")
source("convertDesign.R")
system("zip metawards.zip model_code/* R_tools/* vignette/* convertDesign.R extractOutput.R runscript.sh README.sh")
setwd(tempdir)
```

