---  
title: Model Description
author: "TJ McKinley"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_document:
    highlight: zenburn
    css: css/main.css
---

<center>

**Requires MetaWards 1.2.0**

Please download files for running the model from **[here](metawards.zip)**.

</center>

<a name="model"></a>

The basic model structure is:

* $E$: infected but not infectious;
* $A$: asymptomatic and infectious;
* $I$: symptomatic and infectious;
* $H$: hospitalised and infectious;
* $C$: in critical care (ICU) and infectious;
* $R$: recovered and immune;
* $D$: died.

```{r, echo = FALSE}
library(DiagrammeR)
grViz("digraph simpmod {
    graph[layout = dot, rankdir = LR]
    node [shape = rectangle]
    
    S [group = gp]; E [group = gp]; I [group = gp]; RI [label = 'R@_{I}', group = gp]; DI [label = 'D@_{I}', group = gp]; 
    A [group = as]; RA [label = 'R@_{A}', group = as];
    H [group = hos]; RH [label = 'R@_{H}', group = hos]; DH [label = 'D@_{H}', group = hos]; 
    C [group = crit]; RC [label = 'R@_{C}', group = crit]; DC [label = 'D@_{C}', group = crit];
    genpop [label = 'Gen. Pop.', group = gp]
    asymp [label = 'Asymptomatics', group = as]
    hospital [label = 'Hospital Patient', group = hos]
    critical [label = 'Critical Care', group = crit]
    St00 [style = 'invis', group = st]; St0 [label = 'Stage 0', group = st]; 
    St1 [label = 'Stage 1', group = st]; 
    St2 [label = 'Stage 2', group = st]; St3 [label = 'Stage 3', group = st]; 
    
    genpop -> S [style = 'invis']
    hospital -> H [style = 'invis']
    critical -> C [style = 'invis']
    asymp -> A [style = 'invis']
    S -> E -> I -> RI
    I -> DI [weight = 0]
    I -> H -> C -> RC 
    C -> DC [weight = 0]
    H -> RH 
    H -> DH [weight = 0] 
    RI -> DI [style = 'invis']
    RH -> DH [style = 'invis']
    RC -> DC [style = 'invis']
    E -> A [weight = 0]
    A -> RA 
    St00 -> St0 -> St1 -> St2 -> St3 [style = 'invis']
    subgraph {
      rank = same; St0; E;
    }
    subgraph {
      rank = same; St00 -> asymp -> genpop -> hospital -> critical [style = 'invis']
    }
    subgraph cluster {
      subgraph {
        rank = same; A -> I -> H -> C [style = 'invis']
      }
      style = filled;
      color = lightgrey;
      label = 'Infectious Classes'
    }
    subgraph {
      rank = same; St2; RA, RI, RC; RH;
    }
    subgraph {
      rank = same; St3; DI; DH; DC;
    }
    
}", height = '100%', width = '100%')
```

## MetaWards setup

The model structure above allows for different progression pathways. MetaWards deals with this by assigning individuals to different "demographics", where each demographic can be used to represent different progression pathways. There are also `mover` functions that can be used to move individuals between demographics, and `mixer` functions that scale the force-of-infection (FOI) terms between the different demographics (explained later). There are currently seven pathways, which will be described below, but can be summarised as:

* `SEAR`: asymptomatic infections, always recover.
* `SEIR`: symptomatic infections, leading to recovery.
* `SEID`: symptomatic infections, leading to death.
* `SEIHR`: symptomatic infections, leading to hospital and then recovery.
* `SEIHD`: symptomatic infections, leading to hospital and then death.
* `SEIHCR`: symptomatic infections, leading to hospital, then critical care (ICU) and then recovery.
* `SEIHCD`: symptomatic infections, leading to hospital, then critical care (ICU) and then death.

Individuals in the `genpop` demographic move through the $SEIR$ and $SEID$ pathways; the `asymp` demographic moves through the `SEAR` pathway; some individuals in the `genpop` demographic can be moved to the `hospital` demographic, and thus follow the `SEIHR` and `SEIHD` pathways; and finally, some individuals in the `hospital` demographic can be moved to the `critical` demographic, and thus follow the `SEIHCR` and `SEIHCD` pathways.

We need to specify initial proportions of individuals in each pathway such that $\sum_{i \in \mathcal{P}} p_i = 1$, where $\mathcal{P} = \{GP, A, H, C\}$.

> Here we assume the same ratios in the `work` and `play` populations, but in practice these could be changed.

The demographics in MetaWards is set up using the `demographics.json` file.

<button data-toggle="collapse" data-target="#demographics">Click for MetaWards <code>demographics</code> file</button>
<div id="demographics" class="collapse">

```{js, code = readLines("../model_code/demographics.json"), eval = FALSE}
```

</div>

The number of stages can be set with the MetaWards disease file, called `ncov.json` here.

<button data-toggle="collapse" data-target="#disease">Click for MetaWards <code>disease</code> file</button>
<div id="disease" class="collapse">

All pathways have an $SEI(R/D)$ structure, so `ncov.json` contains:

```{js, code = readLines("../model_code/ncov.json"), eval = FALSE}
```

We will discuss these choices in more detail in the subsequent sections, in particular the role of the `beta` parameters in the next section. Note that the `progress` parameters are all set to zero here, since all movements will be controlled via the custom `mover` function.

</div>

## Parameters {#parameters}

For generic states $X$ and $Y$ say, the progress of an individual from state $X$ to state $Y$ in a given day is governed by probability $q_{XY}$, where:

* $q_{SE}$ driven by $\beta$ parameters, defined by $R_0$ and length of infectious period (see [here](#code));
* $q_{EA} = p_{EA}\left(1 - e^{-\gamma_{E}}\right)$ where $\gamma_E = \frac{1}{T_E}$ with $T_E$ the mean incubation period;
* $q_{EI} = \left(1 - p_{EA}\right)\left(1 - e^{-\gamma_{E}}\right)$;
* $q_{AR} = 1 - e^{-\gamma_I}$ where $\gamma_I = \frac{1}{T_I}$ with $T_I$ the mean infectious period;
* $q_{IH} = p_{IH}\left(1 - e^{-\gamma_{I}}\right)$;
* $q_{IR} = p_{IR}\left(1 - e^{-\gamma_{I}}\right)$;
* $q_{ID} = \left(1 - p_{IH} - p_{IR}\right)\left(1 - e^{-\gamma_{I}}\right)$;
* $q_{HC} = p_{HC}\left(1 - e^{-\gamma_{H}}\right)$ where $\gamma_{H} = \frac{1}{T_H}$ with $T_H$ the mean length of a hospital stay;
* $q_{HR} = p_{HR}\left(1 - e^{-\gamma_{H}}\right)$;
* $q_{HD} = \left(1 - p_{HC} - p_{HR}\right)\left(1 - e^{-\gamma_{H}}\right)$;
* $q_{CR} = p_{CR}\left(1 - e^{-\gamma_{C}}\right)$ where $\gamma_C = \frac{1}{T_C}$ with $T_C$ is the mean length of time spent in ICU;
* $q_{CD} = \left(1 - p_{CR}\right)\left(1 - e^{-\gamma_{C}}\right)$.

Lockdown can be controlled either by constraining movements over a certain distance, or by scaling the $\beta$ parameters. We do the latter at the moment. 

MetaWards has a specific structure in terms of how it progresses movements between the stages. To get the correct splits between the pathways specified above we do all non-infection movements through a custom `mover` function specified [below](#mover). Note that in the `ncov.json` file specified [above](#diseasefile), we set all `progress` parameters to be 0. Thus, all transition probabilities other than new infections are driven by user-defined parameters that are passed to the `mover` function, which we call `move_pathways.py` here.

<button data-toggle="collapse" data-target="#mover">Click for MetaWards <code>mover</code> file</button>
<div id="mover" class="collapse">

The `mover` function applies movements in order, and so it is important to get the order correct. In particular we need to reverse the order of the stage movements (e.g. do movements out of the $C$ demographic *before* movements out of the $H$ demographic). This is to ensure that individuals that move from $H \to C$ say, can't then immediately move out of $C$. The file `move_pathways.py` contains the code below.

```{python, code = readLines("../model_code/move_pathways.py"), eval = FALSE}
```

> **Additional note**: The functions in the `mover` file operate in turn. Therefore the movement probabilities [above](#parameters) must be altered between each function, in order to get the correct proportions moved. For example, consider that we have $n$ individuals in the $I$ class and we want to move a proportion $p_Ip_{IH}$ from $I \to H$, a proportion $p_Ip_{IR}$ from $I \to R_I$, and a proportion $p_I\left(1 - p_{IH} - p_{IR}\right)$ from $I \to D_I$, where $p_I = 1 - e^{-\gamma_I}$.
> 
> In this case the first `mover` function takes a random binomial sample from the $n$ individuals with probability $p_Ip_{IH}$ as requested, resulting in $n_{IH}$ moves. However, the second `mover` function now operates on the $n - n_{IH}$ individuals, so we need to adjust the sampling probabilities to adjust for this. Hence the second `mover` function needs to sample from the $n - n_{IH}$ individuals with probability $\frac{p_Ip_{IR}}{1 - p_Ip_{IH}}$ in order to generate the correct proportions of moves that we would expect, resulting in $n_{IR}$ moves. Similarly, the third `mover` function now operates on the $n - n_{IH} - n_{IR}$ remaining individuals, and thus we would need to adjust the sampling probability to $\frac{p_I\left(1 - p_{IH} - p_{IR}\right)}{1 - p_I\left(p_{IH} + p_{IR}\right)}$. The remaining individuals remain in $I$.

</div>

## Interaction matrices

The **interaction matrix** scales the force-of-infection (FOI) that different demographics have on other demographics. We have three free parameters here: $\beta^S_{A \to GP}$, $\beta^S_{H \to GP}$ and $\beta^S_{C \to GP}$ which scale the impact of aymptomatics, hospital cases and critical care cases on the FOI to the general population. We store this information in a file called `mix_pathways.py`. 

<button data-toggle="collapse" data-target="#mixer">Click for MetaWards <code>mixing</code> file</button>
<div id="mixer" class="collapse">

Since most of the classes are already infected, it's only the first row that has to contain non-zero entries. Here the `GP_A`, `GP_H` and `GP_C` parameters scale the impact of aymptomatics, hospital cases and critical care cases on the force-of-infection to the general population. All demographics other than `genpop` are already infected, so have no further force-of-infection acting on them, and thus the other parameters can be set to zero. These parameters are all $> 0$.

```{python, code = readLines("../model_code/mix_pathways.py"), eval = FALSE}
```

</div>

## Lockdown

Chris Fenton wrote a neat custom `iterator` function, that models lockdown by scaling the FOI terms for different time periods, which each represent a different stage of interventions. We store this information in a custom `iterator` function (specified in `iterator.py` here), with the corresponding parameters in `lockdown_states.txt`.

<button data-toggle="collapse" data-target="#ldown">Click for MetaWards <code>iterator</code> files</button>
<div id="ldown" class="collapse">

The custom iterator is in the `iterate.py` file:

```{python, code = readLines("../model_code/iterate.py"), eval = FALSE}
```

This requires an additional input file containing the dates of the different stages (which are fixed parameters here). This is stored in the `lockdown_states.txt` file:

```{python, code = readLines("../model_code/lockdown_states.txt"), eval = FALSE}
```

</div>

## Extractor {#extractor}

We also have a custom `extractor` function, that saves the outputs as a compressed SQL database called `stages.db.bz2`. This is important since the outputs are very large (all classes across all days and all wards, for multiple replicates of each design point). 

The database for a given run contains a single table called `compact`. To store the outputs efficiently, we have introduced various tricks:

1. Extraneous classes are removed where not relevant for different demographics. For example, for the `genpop` demographic we return `stage_0`--`stage_3`. However, for the `asymp` demographic we only return `stage_1` and `stage_2`, and for the `hospital` and `critical` demographics, we return `stage_1`--`stage_3`.
2. We only return data for time points *after initial infection in a ward*. This also means that wards that are not infected do not return any results. 

These tricks **hugely** reduce the size of the output data, but means that we have to do some **post-processing** in order to extract quantities of interest. Since the data are stored as an SQL database, we can either query the database directly, or use some of the tools in e.g. R (see below) to interface with it. So, to clarify, the `stages.db` database contains a table called `compact` with entries:

* `day`, `ward`
* `Einc`, `E`, `Iinc`, `I`, `RI`, `DI`
* `Ainc`, `A`, `RA`
* `Hinc`, `H`, `RH`, `DH`
* `Cinc`, `C`, `RC`, `DC`

For the `genpop` demographic, these entries correspond to:

* `Einc`: **new** infections on each day (i.e. new moves into $E$)
* `E`: current infections on each day (i.e. number of individuals in $E$)
* `Iinc`: **new** infectious individuals on each day (i.e. new moves into $I$)
* `I`: current infectious individuals on each day (i.e. number of individuals in $I$)
* `RI`: current removals on each day (i.e. number of individuals in $R$)
* `DI`: current deaths on each day (i.e. number of individuals in $D$)

Similarly for the other demographics (refer to the model structure [here](#model)).

We save this custom `extractor` function in the file `ward_extractor.py`.

<button data-toggle="collapse" data-target="#ext">Click for MetaWards <code>extractor</code> file</button>
<div id="ext" class="collapse">

```{python, code = readLines("../model_code/ward_extractor.py"), eval = FALSE}
```

</div>

## Input and output code {#code}

To run designs, we need to generate a `disease.csv` file containing different parameters to use for different runs. For consistency, we will define three spaces:

* *input* space: this relates to the parameters ranges (defined below);
* *design* space: this will usually be in $(0, 1)$ or $(-1, 1)$ space;
* *disease* space: this relates to parameters that are fed into MetaWards.

The *input* and *design* spaces are fairly trivial to convert between, but some more work has to be done to convert between the *input* space and the *disease* space. To ensure orthogonality, we reparameterise a couple of the parameters such that:
\begin{align}
    p_{IR} &= \left(1 - p_{IH}\right) p^\prime_{IR}\\
    p_{HR} &= \left(1 - p_{HC}\right) p^\prime_{HR}
\end{align}
and thus we have parameter ranges:
\begin{align}
    \mbox{$R_0$}&: (2, 4)\\
    \mbox{mean incubation period ($T_E$)}&: (4, 6)\\
    \mbox{mean infectious period ($T_I$)}&: (2, 4)\\
    \mbox{mean hospital stay ($T_H$)}&: (4, 12)\\
    \mbox{mean time in critical care ($T_C$)}&: (4, 12)\\
    \mbox{lockdown 1 restrict}&: (0, 1)\\
    \mbox{lockdown 2 release}&: (0, 1)\\
    p_{EA}&: (0, 1)\\
    p_{IH}&: (0, 1)\\
    p^\prime_{IR}&: (0, 1)\\
    p_{HC}&: (0, 1)\\
    p^\prime_{HR}&: (0, 1)\\
    p_{CR}&: (0, 1)\\
    \beta^S_{A \to GP}&: (0, 1)\\
    \beta^S_{H \to GP}&: (0, 1)\\
    \beta^S_{C \to GP}&: (0, 1).
\end{align}

In R we can set up the *input* parameter ranges as follows:

```{r}
## set up parameter ranges
parRanges <- data.frame(
    parameter = c("r_zero", "incubation_time", "infectious_time", "hospital_time",
                  "critical_time", "lock_1_restrict", "lock_2_release",
                  "pEA", "pIH", "pIRprime", "pHC", "pHRprime", "pCR", 
                  "GP_A", "GP_H", "GP_C"),
    lower = c(2, 4, 2, 4, 4, rep(0, 11)),
    upper = c(4, 6, 4, 12, 12, rep(1, 11)),
    stringsAsFactors = FALSE
)
```

Firstly we want a function to convert between the *design* and *input* spaces. A short R function called `convertDesignToInput()` which does this is given below. This requires a `design` data frame with columns denoting each *input* parameter in `parRanges` and rows corresponding to design points. There should be an additional column called `output` that defines a unique identifier for each design point, and a column called `repeats` that contains the number of repeats for each design point. The `convertDesignToInput()` function also requires the `parRanges` data frame (defined above). We use the `scale` argument to define whether the design is on the $(0, 1)$ (`scale = "zero_one"`) or $(-1, 1)$ (`scale = "negone_one"`) space.

> **Note**: this function converts the probabilities correctly for the `mover` defined [above](#mover).

```{r, echo = FALSE}
library(knitr)
read_chunk("../R_tools/dataTools.R")
```

<button data-toggle="collapse" data-target="#convert">Click for R <code>convertDesignToInput</code> function</button>
<div id="convert" class="collapse">

```{r, convertDesignToInput}
```

</div>

Once we have done this, we need to transform from the *input* space to the *disease* space for MetaWards. A `convertInputToDisease()` R function is given below. This requires an `input` data frame, with columns denoting each *input* parameter and rows corresponding to each input points, a number of `repeats` and a column of unique identifiers (`output`).

<button data-toggle="collapse" data-target="#convert1">Click for R <code>convertInputToDisease</code> function</button>
<div id="convert1" class="collapse">

```{r, convertInputToDisease}
```

</div>

Also in `dataTools.R` is a function `ensembleIDGen()` that creates unique IDs for each design point. So an example of a quick LHS design for five design points and five replicates is:

```{r, code = readLines("../convertDesign.R"), eval = FALSE}
```

This produces a file `inputs/disease.dat` that can be passed to MetaWards to run the model. The `runscript.sh` batch file provides the command line instructions needed to the model using these inputs. This is written for Linux, and should be able to be run directly from the downloaded folder. If you don't run Linux, then the file should give you an idea of how to run the model on your own system.

<button data-toggle="collapse" data-target="#run">Click for <code>runscript.sh</code> file</button>
<div id="run" class="collapse">

```{bash, code = readLines("../runscript.sh"), eval = FALSE}
```

</div>

### Outputs 

As described [above](#extractor) each model run produces a file called `stages.db.bz2`, which is a compressed database containing the outputs. As an example of how to manipulate this, migrate to an output folder containing `stages.db.bz2`. To access the outputs you will first have to unzip this file. I do this on the command line e.g.

```{bash}
bzip2 -dkf stages.db.bz2
```

You will notice that the unzipped `stages.db` file is larger than the compressed version, though Chris has done a great job in storing the outputs in a memory efficient way. As such, you might need to remove `stages.db` at the end if you have limited hard drive space (to this end, the `bzip -dkf` flag I used above ensures that the original compressed file is not deleted when uncompressed). 

The database contains a single table called `compact`. To store the outputs efficiently, we have introduced various tricks, described [above](#extractor). To clarify, the `stages.db` database contains a table called `compact` with entries:

* `day`, `ward`
* `Einc`, `E`, `Iinc`, `I`, `RI`, `DI`
* `Ainc`, `A`, `RA`
* `Hinc`, `H`, `RH`, `DH`
* `Cinc`, `C`, `RC`, `DC`

If you're happy with SQL, you can query these directly with e.g. SQLite. If you are an R user, then the `dplyr` package (or more specifically the `dbplyr` package) provides some useful R tools for querying SQL databases using `tidyverse`-type notation. More details about these tools can be found [here](https://cran.r-project.org/web/packages/dbplyr/vignettes/dbplyr.html).

As a quick example, imagine that we want to extract the **cumulative hospital cases** on say day 100. Here we will need to extract the **new hospital cases** from day 1--100, and then sum them up for each ward. Therefore we need to extract `day`, `ward` and `Hinc` from the `compact` table.

```{r, message = FALSE, warning = FALSE}
## load library
## (you might also need to install the 'RSQLite' 
## and `dbplyr` packages which 'dplyr' calls)
library(dplyr)

## establish connection to database
con <- DBI::dbConnect(RSQLite::SQLite(), "stages.db")

## connect to the 'compact' table
compact <- tbl(con, "compact")

## examine
compact
```

By default, the package only pulls down enough data from the database to produce a summary on the screen (notice that it prints the dimensions as `?? x 19`). If the database is small enough, then the `collect()` function can be used to import tables directly into R as a `tibble`. Alternatively, the `dbplyr` package can convert `tidyverse`-style commands into SQL and run these directly within the database. For large databases this is likely to be much more efficient, in terms of speed and memory usage. You can then `collect()` the results of the query into R.

As an example of this latter approach, we will set up a query that sums the **new** hospital cases over the first 100 days in each ward. **Remember**: for each ward the database only contains days after the initial infection occurs. For cumulative *incidence* counts this is fine, since the missing data will be zero in each case, so we just need to filter first to remove all time points $> 100$. To set up the query:

```{r}
## Hinc contains the new cases, so sum these
## over each ward for days 1--100
hosp_db <- filter(compact, day <= 100) %>%
    select(ward, Hinc) %>%
    group_by(ward) %>%
    summarise(Hcum = sum(Hinc))
```

This hasn't run any commands yet, rather `hosp_db` contains a parsed SQL query that can be run through the database connection. If you like, you can view the generated SQL query using:

```{r}
## view query
show_query(hosp_db)
```

Now let's run the query and return the results to R by passing `hosp_db` to the  `collect()` function:

```{r}
## run query and pull to R
hosp <- collect(hosp_db)

## disconnect from database
DBI::dbDisconnect(con)

## print to screen
hosp
```

Now you can play with `hosp` as much as you like. **Note** that `hosp` here only contains information about wards that have some infections, and only from the time since initial infection. Hence for calibration you might need to expand to fill in the missing wards. Fortunately, R (especially `tidyverse`) has lots of tools for doing this. 

Note that after you've pulled the correct outputs down, you might want to delete the `stages.db` (**NOT** the `stages.db.bz2`) file. I did this on the Unix command line using:

```{bash, eval = TRUE}
rm stages.db
```

> **Be careful**: remember that `rm` removes completely, so not to the recycle bin. However, as long as you don't remove `stages.db.bz2` then you can always recover.

You can very easily wrap these ideas into an R function that can scroll through the design IDs, extract relevant outputs and bind to the inputs. Some examples are described below.

#### SQL outputs

The file `extractOutputsSQL.R` provides code that extracts mean hospital prevalence, mean critical care prevalence, and total deaths from critical care and hospitals during each week since lockdown, for each ward. This output is large, so the code produces a compressed SQL database: `raw_outputs/uberStages.db.bz2`. (This also depends on two other files `extractSQL.sql` and `mergeSQL.sql`, which are both available in the code [zip](./metawards.zip) file.) This does include some OS-specific code that might not work on your machine. Please view this as an illustration, which you might have to amend depending on your own setup.

There is subsequently a file called `extractOutputs.R` that includes some R code giving examples about how to manipulate and extract information from `uberStages.db`. Note that in order to do this you will have to uncompress the `uberStages.db.bz2` file. On Linux this can be done on the command line as above, but bear in mind that for large designs the uncompressed file will be **huge**, and it will take a while to decompress. Whether you choose to store the decompressed file permanently will depend on how much storage you have!

Please feel free to amend these output files as you see fit, or write your own depending on what you want to extract!

```{r, echo = FALSE, message = FALSE, warning = FALSE}
tempdir <- getwd()
setwd("..")
source("convertDesign.R")
system("zip metawards.zip model_code/* R_tools/* vignette/* convertDesign.R extract* mergeSQL.sql runscript.sh README.sh")
setwd(tempdir)
```

